---
title: "Derive a custom reference signature with PhenoMapR"
output:
  html_document:
    # code_folding: hide
    toc: true
---

## Overview
Some rare cancer types are not represented in TCGA or PRECOG and are therefore not available for the built-in reference scoring in **`PhenoMapR`**. Additionally, you might have a proprietary dataset (cancer or not) with bulk gene expression and sample annotations that you would like to use as a phenotype signature reference. For these cases, **`PhenoMapR`** provides functionality to take a gene expression file and a sample annotation file and will generate gene-level z-scores for your phenotype of interest. This is performed with the **`derive_reference_from_bulk()`** function.

Here, we use the [GSE253260](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE253260) dataset to demonstrate how **`PhenoMapR`** generates phenotype signatures for use in downstream **`PhenoMap()`** scoring. GSE253260 contains bulk RNA-seq data for 317 PDAC samples with survival and other clinical annotations. 

This vignette uses the following pre-processed data:

- `GSE253260_expression.rds` (expression matrix, genes × samples)
- `GSE253260_info.rds` (clinical metadata)

1. Load expression and sample annotations and visualize metadata.
2. **Derive a custom gene z-score reference** from expression and overall survival with **`derive_reference_from_bulk()`**.
3. Score samples with the resulting custom reference.
4. Stratify GSE253260 overall survival by its **own** custom signature score (median split).

## 1. Load expression and phenotype data
Load the pre-processed bulk gene expression and sample annotations. This dataset contains a range of patients across different disease stages, and we will get back to this later in the vignette.
```{r load-data and cohort-demographics, fig.width=9, fig.height=3, fig.align='center'}
suppressPackageStartupMessages(library(PhenoMapR))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(dplyr))

# Vignette data: Google Drive
# requireNamespace("googledrive", quietly = TRUE)
googledrive::drive_deauth()
options(googledrive_quiet = TRUE)

# Load expression data
googledrive::drive_download(
  googledrive::as_id("1NXLzKx0O6q-9Bnl_qPwA0hhijIB9jk-6"),
  path = "GSE253260_expression.rds",
  overwrite = TRUE
)

bulk_mat <- readRDS("GSE253260_expression.rds")

# Load sample annotation data
googledrive::drive_download(
  googledrive::as_id("1j0syd8soOQNkudhGnc407TVCJuuvm4jk"),
  path = "GSE253260_info.rds",
  overwrite = TRUE
)

pheno <- readRDS("GSE253260_info.rds")

# Plot cohort demographics
dat_demo <- pheno %>%
  transmute(
    Age = suppressWarnings(as.numeric(Age)),
    Sex = as.character(Sex),
    ECOG = suppressWarnings(as.numeric(ECOG)),
    Stage = suppressWarnings(as.factor(Disease_Stage))
  )

if (nrow(dat_demo) > 0) {
 # Age histogram with mean line and text
mean_age <- mean(subset(dat_demo, is.finite(Age))$Age, na.rm = TRUE)

p_age <- ggplot(subset(dat_demo, is.finite(Age)), aes(x = Age)) +
  geom_histogram(bins = 30, fill = "grey75", color = "white", linewidth = 0.2) +
  geom_vline(xintercept = mean_age, linetype = "dashed", color = "black", linewidth = 0.8) +
  annotate("text", x = mean_age, y = Inf, 
           label = paste0("Mean = ", round(mean_age, 1)), 
           vjust = 1.5, hjust = 1.5, color = "black", size = 3.5) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = NULL, y = "# of Patients", title = "Age")

# Sex bar plot with counts on top
p_sex <- ggplot(dat_demo, aes(x = Sex)) +
  geom_bar(fill = "grey55", color = "white", linewidth = 0.2) +
  geom_text(stat = 'count', aes(label = after_stat(count)), vjust = -0.5, size = 3.5) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = NULL, y = "# of Patients", title = "Sex") +
  ylim(0, max(table(dat_demo$Sex)) * 1.1)  # Add space for labels

# ECOG bar plot with counts on top
dat_ecog <- subset(dat_demo, is.finite(ECOG))
dat_ecog$ECOG <- factor(dat_ecog$ECOG)

p_ecog <- ggplot(dat_ecog, aes(x = ECOG)) +
  geom_bar(fill = "grey35", color = "white", linewidth = 0.2) +
  geom_text(stat = 'count', aes(label = after_stat(count)), vjust = -0.5, size = 3.5, color = "black") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = NULL, y = "# of Patients", title = "ECOG") +
  ylim(0, max(table(dat_ecog$ECOG)) * 1.1)  # Add space for labels

# Stage bar plot with counts on top
dat_stage <- subset(dat_demo, is.finite(Stage))
dat_stage$Stage <- factor(dat_stage$Stage, levels = c("Resectable", "Borderline", "Locally Advanced", "Metastatic"))

p_stage <- ggplot(dat_stage, aes(x = Stage)) +
  geom_bar(fill = "grey15", color = "white", linewidth = 0.2) +
  geom_text(stat = 'count', aes(label = after_stat(count)), vjust = -0.5, size = 3.5, color = "black") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL, y = "# of Patients", title = "Disease Stage") +
  ylim(0, max(table(dat_stage$Stage)) * 1.1)  # Add space for labels
  
  if (requireNamespace("patchwork", quietly = TRUE)) {
    suppressPackageStartupMessages(library(patchwork))
    print(p_age + p_sex + p_ecog + p_stage + patchwork::plot_layout(widths = c(2, 1, 2, 2)))
  } else {
    print(p_age)
    print(p_sex)
    print(p_ecog)
    print(p_stage)
  }
}
```


## 2. Derive a cohort-wide custom reference signature
Before we perform any subset-specific analyses, we will first use all patients in the dataset to derive a survival phenotype z-score signature from the expression data.

Expression for **`derive_reference_from_bulk()`** must be **samples × genes**. For the phenotype dataframe provided, the user needs to specify the sample mapping column, type of phenotype being ussed (in this case it's overall survival), as well as the column names for the phenotype information.
```{r derive-ref, fig.align='center'}
ref_custom <- suppressWarnings(derive_reference_from_bulk(
    bulk_expression = bulk_mat,
    phenotype = pheno,
    sample_id_column = "geo_accession",
    phenotype_type = "survival",
    survival_time = "OS_Time",
    survival_event = "OS_Censor",
    verbose = TRUE
  )
)

# Vizualize the gene z-score signature distribution
PhenoMapR::plot_score_distribution(ref_custom)
```



Each dataset and phenotype will generate unique signatures, with different numbers of genes significantly associated with the phenotype of interest. For GSE253260, the resulting signature heatmap with top genes annotated is as follows:
```{r, fig.height=2, fig.align='center', fig.width=5, fig.align='center'}
suppressPackageStartupMessages({
  library(ComplexHeatmap)
  library(circlize)
  library(grid)
})

# # Using the output of derive_reference_from_bulk() from above
# df_z <- data.frame(gene = rownames(ref_custom), z = as.numeric(ref_custom[[1]]))
# df_z <- df_z[is.finite(df_z$z) & !is.na(df_z$gene) & nzchar(df_z$gene), ]
# df_z <- df_z[order(df_z$z), ]
# df_z$sig <- abs(df_z$z) > 2
# 
# # 1xN matrix, columns ordered by z
# mat <- matrix(df_z$z, nrow = 1)
# colnames(mat) <- df_z$gene
# rownames(mat) <- "Survival z-score"
# 
# # label top/bottom 15 by z
# lab_n <- min(15L, floor(ncol(mat) / 2L))
# mark_at <- c(seq_len(lab_n), (ncol(mat) - lab_n + 1L):ncol(mat))
# mark_labels <- colnames(mat)[mark_at]
# 
# col_fun <- circlize::colorRamp2(
#   c(min(df_z$z), 0, max(df_z$z)),
#   c("#2166AC", "white", "#B2182B")
# )
# 
# # |z|>2 indicator: thin bar on top
# sig_dir <- ifelse(df_z$z <= -2, "neg",
#            ifelse(df_z$z >= 2,  "pos", "none"))
# 
# # Top annotation (marks for gene labels)
# ha_top <- HeatmapAnnotation(
#   marks = anno_mark(
#     at = mark_at,
#     labels = mark_labels,
#     side = "top",
#     labels_gp = gpar(fontsize = 8),
#     link_gp = gpar(col = "grey50", lwd = 0.6),
#     padding = unit(2, "mm")
#   ),
#   which = "column",
#   annotation_height = unit(22, "mm"),
#   show_annotation_name = FALSE
# )
# 
# # Bottom annotation (significance strip)
# ha_bottom <- HeatmapAnnotation(
#   `|z|>2` = anno_simple(
#     sig_dir,
#     col = c(
#       "none" = "transparent",
#       "neg"  = "#2166AC",  # blue for z < -2
#       "pos"  = "#B2182B"   # red  for z > 2
#     ),
#     height = unit(3, "mm"),  # Made taller for visibility
#     border = FALSE
#   ),
#   which = "column",
#   show_annotation_name = TRUE,  # Show the "|z|>2" label
#   annotation_name_side = "right"
# )
# 
# # Create heatmap with BOTH annotations
# ht <- Heatmap(
#   mat,
#   col = col_fun,
#   cluster_rows = FALSE,
#   cluster_columns = FALSE,
#   show_row_names = FALSE,
#   show_column_names = FALSE,
#   top_annotation = ha_top,
#   bottom_annotation = ha_bottom,  # Add bottom annotation
#   heatmap_legend_param = list(
#     title = "Survival z-score",
#     direction = "horizontal",  # Make legend horizontal
#     title_position = "topcenter"
#   )
# )
# 
# # Draw with legend at bottom
# draw(ht, heatmap_legend_side = "bottom")

plot_reference_signature(ref_custom, row_title = "Survival z-score")

```

## 3. Score samples with the cohort-wide custom reference
Now that we have our custom phenotype signature for GSE253260, we use this signature to score all samples in the dataset.
```{r score-custom}
bulk_genes_rows <- t(bulk_mat)

scores_custom <- PhenoMap(
  expression = bulk_genes_rows,
  reference = ref_custom,
  z_score_cutoff = 2,
  verbose = TRUE
)

col_custom <- grep("survival_z|weighted_sum", colnames(scores_custom), value = TRUE)[1]

pheno$score_custom <- scores_custom[match(pheno$geo_accession, rownames(scores_custom)), col_custom]
```


Next, we plot the 
```{r score-distribution, fig.width=9, fig.height=3, fig.align='center'}
suppressPackageStartupMessages(library(ggplot2))

dat_sc <- pheno[is.finite(pheno$score_custom) & !is.na(pheno$score_custom), , drop = FALSE]
if (nrow(dat_sc) > 0) {
  med_sc <- median(dat_sc$score_custom, na.rm = TRUE)
  dat_sc$score_grp <- factor(ifelse(dat_sc$score_custom >= med_sc, "High", "Low"), levels = c("Low", "High"))
  pal_km <- c("Low" = "#2166AC", "High" = "#B2182B")  # Low (favorable), High (adverse)

  p_hist <- ggplot(dat_sc, aes(x = score_custom)) +
    geom_histogram(aes(fill = score_grp), bins = 40, color = "white", linewidth = 0.2) +
    geom_density(aes(y = after_stat(count), color = score_grp), linewidth = 0.7) +
    scale_fill_manual(values = pal_km, drop = FALSE) +
    scale_color_manual(values = pal_km, drop = FALSE) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "black", linewidth = 0.8) +
  # annotate("text", x = 0, y = Inf, 
  #          label = "Favorable score", 
  #          vjust = 1.5, hjust = 1.5, color = "black", size = 3.5) +
  #     annotate("text", x = 0, y = Inf, 
  #          label = "Favorable score", 
  #          vjust = 1.5, hjust = 1.5, color = "black", size = 3.5) +
    theme_minimal() +
    labs(
      x = "Cohort-wide custom signature score",
      y = "Number of samples",
      title = "Score distribution",
      fill = "Median split",
      color = "Median split"
    )

  dat_sc <- dat_sc[order(dat_sc$score_custom), , drop = FALSE]
  dat_sc$rank <- seq_len(nrow(dat_sc))
  p_rank <- ggplot(dat_sc, aes(x = rank, y = score_custom)) +
    geom_point(aes(color = score_grp), size = 1.2, alpha = 0.9) +
    scale_color_manual(values = pal_km, drop = FALSE) +
    theme_minimal() +
    labs(
      x = "Samples (ordered by score)",
      y = "Cohort-wide custom signature score",
      title = "Scores across samples",
      color = "Median split"
    )

  if (requireNamespace("patchwork", quietly = TRUE)) {
    suppressPackageStartupMessages(library(patchwork))
    print(p_hist + p_rank + patchwork::plot_layout(widths = c(1.2, 1)))
  } else {
    print(p_hist)
    print(p_rank)
  }
}
```

## 4. Survival anlaysis by PhenoMapR score assignment
PhenoMapR scores samples based on the weighted-sum of a sample's gene expression levels and the weights from the phenotype reference signature. Because the signature used to score samples in the dataset was derived from the expression associations with survival across the datasets, the PhenoMapR score should stratify outcomes. We perform Kaplan–Meier analysis using a PhenoMapR median split.

```{r km-custom, fig.width=5, fig.height=4, fig.align='center'}
suppressPackageStartupMessages(library(survival))
pheno$survival_time <- pheno$OS_Time
pheno$survival_event <- pheno$OS_Censor
dat_os <- subset(pheno, !is.na(survival_time) & !is.na(survival_event) & !is.na(score_custom))
if (nrow(dat_os) >= 10) {
  # Convert to months for plotting
  time_months <- dat_os$survival_time
  if (is.finite(max(time_months, na.rm = TRUE)) && max(time_months, na.rm = TRUE) > 400) {
    time_months <- time_months / 30.4375
  }
  dat_os$time_months <- time_months

  dat_os$custom_grp <- factor(
    ifelse(dat_os$score_custom >= median(dat_os$score_custom, na.rm = TRUE), "High", "Low"),
    levels = c("Low", "High")
  )
  fit <- survfit(Surv(time_months, survival_event) ~ custom_grp, data = dat_os)
  lr <- survdiff(Surv(time_months, survival_event) ~ custom_grp, data = dat_os)
  p_lr <- 1 - pchisq(lr$chisq, df = 1)
  cox <- coxph(Surv(time_months, survival_event) ~ custom_grp, data = dat_os)
  hr <- exp(coef(cox))[1]
  ci <- as.vector(exp(confint(cox)))[1:2]
  label <- sprintf("p = %s\nHR = %.2f (%.2f-%.2f)",
                   format.pval(p_lr, digits = 2, eps = 0.001), hr, ci[1], ci[2])
  max_t <- max(dat_os$time_months, na.rm = TRUE)

  if (requireNamespace("survminer", quietly = TRUE)) {
    suppressPackageStartupMessages(library(survminer))
    pal_km <- c("Low" = "#2166AC", "High" = "#B2182B")  # Low (favorable), High (adverse)
    p <- ggsurvplot(
      fit,
      data = dat_os,
      title = NULL,
      xlab = "Time (months)",
      ylab = "Survival probability",
      legend.title = "Custom score",
      legend.labs = c("Low", "High"),
      palette = unname(pal_km[c("Low", "High")]),
      risk.table = FALSE,
      pval = label,
      pval.coord = c(max_t * 0.65, 0.95),
      pval.size = 3.5
    )
    print(p)
  } else {
    plot(fit, col = c("blue", "red"), lwd = 2, xlab = "Time (months)", ylab = "Survival probability",
         main = "GSE253260: OS by custom survival-derived score (median split)")
    legend("bottomleft", legend = c("Low", "High"), col = c("blue", "red"), lwd = 2, bty = "n")
    mtext(label, side = 3, adj = 1, line = -1, cex = 0.8)
  }
}
```
The sample PhenoMapR scores successfully stratify outcomes across the cohort.


## 5. Cohort-wide custom score distribution by disease stage
As mentioned earlier (section 1), this dataset contains patients with a range of disease stages, reflecting significant differences in prognostic risk that could influence the PhenoMapR score assignment. To check if patients with a more adverse PhenoMapR score are enriched in patients with more aggressive disease, we plot the PhenoMapR score distribution across the four disease stage groups.
```{r custom-score-hist, fig.width=7, fig.height=3, fig.align='center'}
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(ggsignif))

stage_map <- function(x) {
  x_low <- tolower(x)
  dplyr::case_when(
    grepl("meta", x_low) ~ "Metastatic",
    grepl("loc", x_low)  ~ "Locally Advanced",
    grepl("border", x_low) ~ "Borderline",
    grepl("res", x_low)  ~ "Resectable",
    TRUE ~ NA_character_
  )
}

pal_stage <- c(
  "Resectable" = "#74B9FF",
  "Borderline" = "#26A69A",
  "Locally Advanced" = "#F39C12",
  "Metastatic" = "#FF6B6B"
)

dat_hist <- dat_os %>%
  mutate(
    geo_accession = as.character(geo_accession),
    stage_group = factor(stage_map(pheno$Disease_Stage[match(geo_accession, pheno$geo_accession)]),
                         levels = names(pal_stage))
  ) %>%
  filter(!is.na(stage_group))

if (nrow(dat_hist) > 0) {
  # ggplot(dat_hist, aes(x = stage_group, y = score_custom, fill = stage_group)) +
  #   geom_violin(trim = FALSE, alpha = 0.85, color = NA) +
  #   geom_boxplot(width = 0.15, outlier.size = 0.4, alpha = 0.9) +
  #   scale_fill_manual(values = pal_stage, drop = FALSE) +
  #   theme_minimal() +
  #   theme(legend.position = "none") +
  #   labs(
  #     x = "Disease stage",
  #     y = "Cohort-wide custom signature score"
  #   )
  library(ggsignif)

ggplot(dat_hist, aes(x = stage_group, y = score_custom, fill = stage_group)) +
  geom_violin(trim = FALSE, alpha = 0.85, color = NA) +
  geom_boxplot(width = 0.15, outlier.size = 0.4, alpha = 0.9) +
  scale_fill_manual(values = pal_stage, drop = FALSE) +
  geom_signif(
    comparisons = combn(levels(dat_hist$stage_group), 2, simplify = FALSE),
    test = "wilcox.test",
    map_signif_level = TRUE,  # TRUE converts p to *, **, ***
    textsize = 3,
    step_increase = 0.08,  # Vertical spacing between brackets
    tip_length = 0.01
  ) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    x = "Disease stage",
    y = "Cohort-wide custom signature score"
  )
}
```
Interestingly, it does not appear that any of the disease risk groups are enriched for favorable or adverse PhenoMapR scores. This suggests that PhenoMapR identifies potentially stage agnostic prognostic signals and might be able to refine the risk within each stage.


## 6. Overall survival by disease stage
Before we test whether the PhenoMapR score can stratify outcomes within each stage, we first confirm that the four different disease stages have significantly different overall survival results.
```{r km-by-stage, fig.width=7, fig.height=3, fig.align='center'}
suppressPackageStartupMessages(library(survival))
suppressPackageStartupMessages(library(dplyr))

stage_map <- function(x) {
  x_low <- tolower(x)
  dplyr::case_when(
    grepl("meta", x_low) ~ "Metastatic",
    grepl("loc", x_low)  ~ "Locally Advanced",
    grepl("border", x_low) ~ "Borderline",
    grepl("res", x_low)  ~ "Resectable",
    TRUE ~ NA_character_
  )
}

pal_stage <- c(
  "Resectable" = "#74B9FF",
  "Borderline" = "#26A69A",
  "Locally Advanced" = "#F39C12",
  "Metastatic" = "#FF6B6B"
)

dat_stage <- data.frame(
  geo_accession = as.character(pheno$geo_accession),
  stage_group = factor(stage_map(pheno$Disease_Stage), levels = names(pal_stage)),
  OS_time_raw = suppressWarnings(as.numeric(pheno$OS_Time)),
  OS_event_raw = suppressWarnings(as.integer(pheno$OS_Censor)),
  stringsAsFactors = FALSE
)

# Heuristic: if time looks like days, convert to months for plotting (as in example figure)
time_months <- dat_stage$OS_time_raw
if (is.finite(max(time_months, na.rm = TRUE)) && max(time_months, na.rm = TRUE) > 400) {
  time_months <- time_months / 30.4375
}
dat_stage$time_months <- time_months

dat_stage <- subset(dat_stage, !is.na(stage_group) & !is.na(time_months) & !is.na(OS_event_raw))
dat_stage$stage_group <- droplevels(dat_stage$stage_group)

if (nrow(dat_stage) >= 10 && nlevels(dat_stage$stage_group) >= 2) {
  fit <- survfit(Surv(time_months, OS_event_raw) ~ stage_group, data = dat_stage)
  lr <- survdiff(Surv(time_months, OS_event_raw) ~ stage_group, data = dat_stage)
  p_lr <- 1 - pchisq(lr$chisq, df = (nlevels(dat_stage$stage_group) - 1))

  cox <- coxph(Surv(time_months, OS_event_raw) ~ stage_group, data = dat_stage)
  hr <- exp(coef(cox))
  ci <- exp(confint(cox))
  hr_lines <- paste0(
    gsub("^stage_group", "", names(hr)),
    ": HR ", sprintf("%.2f", hr),
    " (", sprintf("%.2f", ci[, 1]), "-", sprintf("%.2f", ci[, 2]), ")"
  )
  label <- paste0(
    "p = ", format.pval(p_lr, digits = 2, eps = 0.001),
    "\n", paste(hr_lines, collapse = "\n")
  )
  max_t <- max(dat_stage$time_months, na.rm = TRUE)

  if (requireNamespace("survminer", quietly = TRUE)) {
    suppressPackageStartupMessages(library(survminer))
    pal_vec <- unname(pal_stage[levels(dat_stage$stage_group)])
    p <- ggsurvplot(
      fit,
      data = dat_stage,
      palette = pal_vec,
      risk.table = FALSE,
      legend.title = NULL,
      legend.labs = levels(dat_stage$stage_group),
      xlab = "Time (months)",
      ylab = "Overall survival probability",
      pval = label,
      pval.coord = c(max_t * 0.55, 0.95),
      pval.size = 4
    )
    print(p)
  } else {
    plot(fit, col = pal_stage[levels(dat_stage$stage_group)], lwd = 2,
         xlab = "Time (months)", ylab = "Overall survival probability")
    legend("topright", legend = levels(dat_stage$stage_group),
           col = pal_stage[levels(dat_stage$stage_group)], lwd = 2, bty = "n")
    mtext(paste0("p = ", format.pval(p_lr, digits = 2, eps = 0.001)), side = 4, adj = 1, line = -1, cex = 0.8)
  }
}
```

## 7. Stage-specific custom references and within-stage re-stratification

```{r stage-specific-custom, fig.width=10, fig.height=8}
if (!requireNamespace("survminer", quietly = TRUE)) {
  message("Package 'survminer' not installed; skipping stage-specific KM plots.")
} else if (!requireNamespace("patchwork", quietly = TRUE)) {
  message("Package 'patchwork' not installed; skipping stage-specific KM plots.")
} else {
  suppressPackageStartupMessages(library(survminer))
  suppressPackageStartupMessages(library(patchwork))

  plots <- list()
  stage_levels <- names(pal_stage)

  for (st in stage_levels) {
    dat_st <- subset(dat_stage, stage_group == st)
    if (nrow(dat_st) < 10) next

    # Align expression to these samples
    keep_ids <- intersect(rownames(bulk_mat), dat_st$geo_accession)
    if (length(keep_ids) < 10) next

    bulk_st <- bulk_mat[keep_ids, , drop = FALSE] # samples × genes
    pheno_st <- data.frame(
      sample_id = keep_ids,
      survival_time = dat_st$time_months[match(keep_ids, dat_st$geo_accession)],
      survival_event = dat_st$OS_event_raw[match(keep_ids, dat_st$geo_accession)],
      stringsAsFactors = FALSE
    )
    pheno_st <- pheno_st[!is.na(pheno_st$survival_time) & !is.na(pheno_st$survival_event), , drop = FALSE]
    bulk_st <- bulk_st[pheno_st$sample_id, , drop = FALSE]
    if (nrow(pheno_st) < 10) next

    ref_st <- suppressWarnings(
      derive_reference_from_bulk(
        bulk_expression = bulk_st,
        phenotype = pheno_st,
        sample_id_column = "sample_id",
        phenotype_type = "survival",
        survival_time = "survival_time",
        survival_event = "survival_event",
        gene_axis = "cols",
        verbose = FALSE
      )
    )

    # Score only within this stage using its own signature
    scores_st <- PhenoMap(
      expression = t(bulk_st), # genes × samples
      reference = ref_st,
      z_score_cutoff = 2,
      verbose = FALSE
    )
    col_sc <- grep("survival_z|weighted_sum", colnames(scores_st), value = TRUE)[1]
    sc <- scores_st[, col_sc]

    dat_km <- pheno_st
    dat_km$score_stage <- sc[match(dat_km$sample_id, rownames(scores_st))]
    dat_km <- dat_km[!is.na(dat_km$score_stage), , drop = FALSE]
    if (nrow(dat_km) < 10) next

    med <- median(dat_km$score_stage, na.rm = TRUE)
    dat_km$grp <- factor(ifelse(dat_km$score_stage >= med, "High", "Low"), levels = c("Low", "High"))
    fit <- survfit(Surv(survival_time, survival_event) ~ grp, data = dat_km)
    lr <- survdiff(Surv(survival_time, survival_event) ~ grp, data = dat_km)
    p_lr <- 1 - pchisq(lr$chisq, df = 1)
    cox <- coxph(Surv(survival_time, survival_event) ~ grp, data = dat_km)
    hr <- exp(coef(cox))[1]
    ci <- as.vector(exp(confint(cox)))[1:2]
    label <- sprintf("p = %s\nHR = %.2f (%.2f-%.2f)",
                     format.pval(p_lr, digits = 2, eps = 0.001), hr, ci[1], ci[2])
    max_t <- max(dat_km$survival_time, na.rm = TRUE)

    # Stage color for High; lighter shade for Low
    col_high <- pal_stage[[st]]
    col_low <- grDevices::adjustcolor(col_high, alpha.f = 0.55)

    p <- ggsurvplot(
      fit,
      data = dat_km,
      palette = unname(c(col_low, col_high)),
      risk.table = FALSE,
      legend.title = "Signature score",
      legend.labs = c("Low", "High"),
      title = st,
      xlab = "Time (months)",
      ylab = "Overall survival probability",
      pval = label,
      pval.coord = c(max_t * 0.55, 0.95),
      pval.size = 5
    )

    # Differentiate High vs Low with linetype so we can keep the stage color
    p$plot <- p$plot + ggplot2::aes(linetype = grp) +
      ggplot2::scale_linetype_manual(values = c("Low" = "solid", "High" = "dashed")) +
      ggplot2::guides(color = "none")

    plots[[st]] <- p$plot
  }

  if (length(plots) == 0) {
    message("Not enough samples per stage with OS to derive and plot stage-specific signatures.")
  } else {
    ordered <- plots[intersect(stage_levels, names(plots))]
    print(wrap_plots(ordered, ncol = 2))
  }
}
```

## 8 Conclusions

## 9. References

- Nicole et al Predictive genomic and transcriptomic analysis on endoscopic ultrasound-guided fine needle aspiration materials from primary pancreatic adenocarcinoma: a prospective multicentre study **GSE253260**: [GEO: GSE253260](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE253260).

## Session Info

```{r session-info}
sessionInfo()
```
