---
title: "Scoring single-cell data with PhenoMapR"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
---

## Overview

This vignette demonstrates how to use PhenoMapR on a single-cell RNA-seq dataset. We use the **CRA001160** PDAC dataset from [Peng et al. (*Cell Research* 2019)](https://doi.org/10.1038/s41422-019-0195-y) (single-cell RNA-seq of 57,530 pancreatic cells from primary PDAC tumors and control pancreases; [GSA: CRA001160](https://ngdc.cncb.ac.cn/gsa/browse/CRA001160)). Pre-processed expression and metadata were obtained from [TISCH2](https://tisch.compbio.cn/home/). We score the expression matrix with **TCGA** and **PRECOG** meta-z signatures (no Seurat required), then use the metadata and score outputs for visualizations, prognostic groups, and marker genes. Results are compared with [Jolasun et al.](https://www.nature.com/articles/s41467-025-66162-4) [**SIDISH**](https://www.nature.com/articles/s41467-025-66162-4).

## Load data for CRA001160

Download the expression matrix (10X H5) and cell metadata (TSV) from Google Drive. Align metadata to matrix columns and subsample for vignette runtime.

```{r load-setup}
suppressPackageStartupMessages(library(PhenoMapR))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(googledrive))
if (requireNamespace("ggpubr", quietly = TRUE)) suppressPackageStartupMessages(library(ggpubr))
if (requireNamespace("Seurat", quietly = TRUE)) suppressPackageStartupMessages(library(Seurat))

knitr::opts_chunk$set(fig.width = 12, out.width = "100%")

options(googledrive_quiet = TRUE)
googledrive::drive_deauth()
googledrive::drive_download(googledrive::as_id("1PolTXggREz8XmhutCLTQJGCfKxFAzqMl"), "PAAD_CRA001160_expression.h5", overwrite = TRUE)
googledrive::drive_download(googledrive::as_id("17mqxnKOZJn0jW2iD9RV0wZeWsilAIwdu"), "PAAD_CRA001160_CellMetainfo_table.tsv", overwrite = TRUE)

expr_mat <- Seurat::Read10X_h5("PAAD_CRA001160_expression.h5")
meta <- read.delim("PAAD_CRA001160_CellMetainfo_table.tsv", stringsAsFactors = FALSE, check.names = FALSE)

cell_ids <- colnames(expr_mat)
id_col <- NULL
for (cand in c("Barcode", "barcode", "cell_id", "Cell", "cell_barcode", names(meta)[1])) {
  if (cand %in% names(meta) && length(intersect(meta[[cand]], cell_ids)) > 0) {
    id_col <- cand
    break
  }
}
if (is.null(id_col)) id_col <- names(meta)[1]
meta <- meta[meta[[id_col]] %in% cell_ids, , drop = FALSE]
meta <- meta[match(cell_ids, meta[[id_col]]), , drop = FALSE]
rownames(meta) <- colnames(expr_mat)

for (col in names(meta)) {
  if (!(is.character(meta[[col]]) || is.factor(meta[[col]]))) next
  nlev <- length(unique(meta[[col]][!is.na(meta[[col]])]))
  if (nlev >= 2L && nlev <= 500L) {
    celltype_col <- col
    break
  }
}
if (!exists("celltype_col")) celltype_col <- names(meta)[2]
celltype_original_col <- if ("Celltype (original)" %in% names(meta)) "Celltype (original)" else NULL

max_cells <- 25000L
if (ncol(expr_mat) > max_cells) {
  set.seed(1)
  keep_cells <- sample(colnames(expr_mat), max_cells)
  expr_mat <- expr_mat[, keep_cells, drop = FALSE]
  meta <- meta[keep_cells, , drop = FALSE]
  message(sprintf("Subsampled to %d cells for vignette.", ncol(expr_mat)))
}
message(sprintf("Cells: %d | Genes: %d | Cell type: %s", ncol(expr_mat), nrow(expr_mat), celltype_col))
```

## Score cells with TCGA and PRECOG Pancreatic

Score the expression matrix with both references; merge score columns into metadata.

```{r score-cells}
scores_tcga <- PhenoMap(expression = expr_mat, reference = "tcga", cancer_type = "PAAD", verbose = TRUE)
scores_precog <- PhenoMap(expression = expr_mat, reference = "precog", cancer_type = "Pancreatic", verbose = TRUE)
for (col in names(scores_tcga)) meta[[col]] <- scores_tcga[rownames(meta), col]
for (col in names(scores_precog)) meta[[col]] <- scores_precog[rownames(meta), col]

score_tcga_col <- grep("weighted_sum_score.*PAAD", names(meta), value = TRUE, ignore.case = TRUE)[1]
score_precog_col <- grep("weighted_sum_score.*Pancreatic", names(meta), value = TRUE, ignore.case = TRUE)[1]
if (is.na(score_tcga_col)) score_tcga_col <- names(scores_tcga)[1]
if (is.na(score_precog_col)) score_precog_col <- names(scores_precog)[1]
```

## Scatterplot: TCGA vs PRECOG scores

Comparison of per-cell scores; points colored by cell type with regression line, 95% CI, and correlation by group.

```{r scatter-tcga-precog}
df_scatter <- data.frame(
  x = meta[[score_precog_col]],
  y = meta[[score_tcga_col]],
  Cell_type = factor(meta[[celltype_col]])
)
pal_ct <- PhenoMapR::get_celltype_palette(levels(df_scatter$Cell_type))
if (requireNamespace("ggpubr", quietly = TRUE)) {
  p <- ggpubr::ggscatter(df_scatter, x = "x", y = "y", color = "Cell_type", palette = pal_ct,
    add = "reg.line", conf.int = TRUE, size = 0.5, alpha = 0.3) +
    ggpubr::stat_cor(aes(color = Cell_type), label.x.npc = "left", size = 3.5) +
    scale_x_continuous("PRECOG Pancreatic score") +
    scale_y_continuous("TCGA PAAD score") +
    labs(title = "CRA001160: TCGA vs PRECOG scores by cell type") +
    theme_minimal(base_size = 14) +
    theme(legend.position = "right", legend.title = element_text(size = 12), legend.text = element_text(size = 10),
      axis.title = element_text(size = 14), axis.text = element_text(size = 12))
  print(p)
} else {
  print(ggplot(df_scatter, aes(x = x, y = y, color = Cell_type)) +
    geom_point(alpha = 0.3, size = 0.5) +
    geom_smooth(method = "lm", se = TRUE, linewidth = 0.5) +
    scale_color_manual(values = pal_ct, name = "Cell type") +
    theme_minimal(base_size = 14) +
    labs(x = "PRECOG Pancreatic score", y = "TCGA PAAD score", title = "CRA001160: TCGA vs PRECOG scores"))
}
```

PhenoMap scores are correlated between TCGA and PRECOG across cell types.

## Score distribution: TCGA and PRECOG by cell type (ordered by PRECOG)

Boxplots of score distributions for both references, with cell types ordered by median PRECOG score so that the most adverse (high PRECOG) cell types appear toward one side.

```{r score-distribution-boxplots}
med_precog <- setNames(
  tapply(meta[[score_precog_col]], meta[[celltype_col]], median, na.rm = TRUE),
  levels(factor(meta[[celltype_col]]))
)
med_precog <- med_precog[!is.na(med_precog)]
ct_order <- names(sort(med_precog))
meta[[celltype_col]] <- factor(meta[[celltype_col]], levels = ct_order)

dl <- rbind(
  data.frame(Reference = "TCGA PAAD", Score = meta[[score_tcga_col]], Cell_type = meta[[celltype_col]], stringsAsFactors = FALSE),
  data.frame(Reference = "PRECOG Pancreatic", Score = meta[[score_precog_col]], Cell_type = meta[[celltype_col]], stringsAsFactors = FALSE)
)
dl$Reference <- factor(dl$Reference, levels = c("TCGA PAAD", "PRECOG Pancreatic"))
pal_ct <- PhenoMapR::get_celltype_palette(ct_order)
ggplot(dl, aes(x = Cell_type, y = Score, fill = Cell_type)) +
  geom_boxplot(outlier.alpha = 0.2) +
  facet_wrap(~ Reference, ncol = 2, scales = "free_y") +
  scale_fill_manual(values = pal_ct, name = "Cell type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "right", strip.text = element_text(size = 12)) +
  labs(x = "Cell type (ordered by PRECOG median)", y = "PhenoMapR score", title = "CRA001160: Score distribution by reference and cell type")
```

PhenoMapR successfully identifies the Malignant cell compartment as the most associated with the adverse prognostic signature in both TCGA and PRECOG.

## Score distribution by refined cell type

To see if the PhenoMapR score assignment is enriched in more granular cell types, we use the original cell type labels provided by the authors (*Peng et al.* 2019).

```{r score-distribution-celltype-original}
if (!is.null(celltype_original_col) && celltype_original_col %in% names(meta)) {
  med_precog_orig <- setNames(
    tapply(meta[[score_precog_col]], meta[[celltype_original_col]], median, na.rm = TRUE),
    levels(factor(meta[[celltype_original_col]]))
  )
  med_precog_orig <- med_precog_orig[!is.na(med_precog_orig)]
  ct_order_orig <- names(sort(med_precog_orig))
  meta$celltype_original <- factor(meta[[celltype_original_col]], levels = ct_order_orig)
  dl_orig <- rbind(
    data.frame(Reference = "TCGA PAAD", Score = meta[[score_tcga_col]], Cell_type = meta$celltype_original, stringsAsFactors = FALSE),
    data.frame(Reference = "PRECOG Pancreatic", Score = meta[[score_precog_col]], Cell_type = meta$celltype_original, stringsAsFactors = FALSE)
  )
  dl_orig$Reference <- factor(dl_orig$Reference, levels = c("TCGA PAAD", "PRECOG Pancreatic"))
  pal_ct_orig <- PhenoMapR::get_celltype_palette(ct_order_orig)
  print(ggplot(dl_orig, aes(x = Cell_type, y = Score, fill = Cell_type)) +
    geom_boxplot(outlier.alpha = 0.2) +
    facet_wrap(~ Reference, ncol = 2, scales = "free_y") +
    scale_fill_manual(values = pal_ct_orig, name = "Celltype (original)") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "right", strip.text = element_text(size = 12)) +
    labs(x = "Celltype (original)", y = "PhenoMapR score", title = "CRA001160: Score by reference and Celltype (original)"))
} else {
  message("Celltype (original) not in metadata; skipping boxplots by original annotations.")
}
```

In both TCGA and PRECOG results, we see that the Ductal cell type 2 is the most significantly associated cell type with an adverse prognostic signature in PAAD. This agrees with the results from SIDISH [Jolasun et al.](https://www.nature.com/articles/s41467-025-66162-4/figures/2), where they find over 55% of their high-risk cells were Ductal cell type 2. 

## Cell type representation in the most phenotypically associated populations

Cell type counts in the top 5% (Most Adverse) and bottom 5% (Most Favorable) of cells by PRECOG score.

```{r celltype-in-extreme-percentiles}
q_lo <- quantile(meta[[score_precog_col]], 0.05, na.rm = TRUE)
q_hi <- quantile(meta[[score_precog_col]], 0.95, na.rm = TRUE)
pg_tmp <- ifelse(meta[[score_precog_col]] >= q_hi, "Most Adverse", ifelse(meta[[score_precog_col]] <= q_lo, "Most Favorable", NA))
idx_extreme <- !is.na(pg_tmp)
ct_in_extreme <- meta[idx_extreme, ]
ct_in_extreme$pg <- pg_tmp[idx_extreme]
ct_counts <- as.data.frame(table(Cell_type = ct_in_extreme[[celltype_col]], pg = ct_in_extreme$pg))
ct_wide <- reshape(ct_counts, idvar = "Cell_type", timevar = "pg", direction = "wide")
names(ct_wide) <- gsub("Freq\\.", "", names(ct_wide))
for (c in c("Most Adverse", "Most Favorable")) {
  if (!c %in% names(ct_wide)) ct_wide[[c]] <- 0
}
ct_wide$Cell_type <- factor(ct_wide$Cell_type, levels = ct_wide$Cell_type[order(ct_wide$`Most Adverse`, decreasing = TRUE)])
ct_long <- rbind(
  data.frame(Cell_type = ct_wide$Cell_type, pg = "Most Favorable", n = -ct_wide$`Most Favorable`),
  data.frame(Cell_type = ct_wide$Cell_type, pg = "Most Adverse", n = ct_wide$`Most Adverse`)
)
ct_long$pg <- factor(ct_long$pg, levels = c("Most Favorable", "Most Adverse"))
max_n <- max(abs(ct_long$n), 1)
ggplot(ct_long, aes(x = Cell_type, y = n, fill = pg)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c(`Most Adverse` = "#B2182B", `Most Favorable` = "#2166AC"), name = "Prognostic group") +
  scale_y_continuous(breaks = seq(-max_n, max_n, length.out = 5), labels = abs(seq(-max_n, max_n, length.out = 5))) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 9), legend.position = "right") +
  labs(x = "Cell type", y = "Cell count (5th %ile)", title = "CRA001160: Cell type in adverse vs favorable extremes")
```

## Prognostic groups and marker genes

We define the "Most Adverse" and "Most Favorable" cells from the **PRECOG** score (5th/95th percentiles), then run marker identification (adverse vs rest, favorable vs rest) using the original cell type annotation for context.

```{r prognostic-groups-markers, cache = FALSE}
scores_df <- meta[, c(score_tcga_col, score_precog_col), drop = FALSE]
groups <- define_prognostic_groups(scores_df, percentile = 0.05, score_columns = score_precog_col)
group_col <- grep("prognostic_group", names(groups), value = TRUE)[1]
meta[[group_col]] <- groups[rownames(meta), group_col]

markers <- NULL
if (!is.na(group_col)) {
  markers <- find_prognostic_markers(expr_mat, group_labels = meta[[group_col]], max_cells_per_ident = 5000L)
  if (!is.null(markers)) {
    message("Adverse markers (top 5):"); print(head(markers$adverse_markers, 5))
    message("Favorable markers (top 5):"); print(head(markers$favorable_markers, 5))
  }
}
```

## Marker heatmap

Top adverse and favorable marker genes; columns ordered by PhenoMapR score. Annotations: PhenoMapR score (blue &lt; 0, red &gt; 0), cell type, sample type (healthy/tumor), prognostic group.

```{r heatmap-markers}
if (is.null(markers)) {
  message("Markers not available; skipping heatmap.")
} else {
  n_top <- 15
  adverse_pos <- markers$adverse_markers[markers$adverse_markers$avg_log2FC > 0, ]
  favorable_pos <- markers$favorable_markers[markers$favorable_markers$avg_log2FC > 0, ]
  top_genes <- unique(c(
    head(adverse_pos$gene[order(adverse_pos$p_adj)], n_top),
    head(favorable_pos$gene[order(favorable_pos$p_adj)], n_top)
  ))
  top_genes <- top_genes[top_genes %in% rownames(expr_mat)]
  if (length(top_genes) == 0) top_genes <- head(rownames(expr_mat), 20)

  mat <- as.matrix(expr_mat[top_genes, , drop = FALSE])
  mat_scaled <- t(scale(t(mat)))
  mat_scaled[mat_scaled < -3] <- -3
  mat_scaled[mat_scaled > 3] <- 3

  ord <- order(meta[[score_precog_col]])
  mat_scaled <- mat_scaled[, ord, drop = FALSE]
  meta_ord <- meta[ord, ]

  score_vals <- meta_ord[[score_precog_col]]
  lim <- max(abs(range(score_vals, na.rm = TRUE)), 0.01)
  score_norm <- score_vals / lim
  ct_anno_col <- if (!is.null(celltype_original_col) && celltype_original_col %in% names(meta_ord)) celltype_original_col else celltype_col
  sample_type_col <- NULL
  for (cand in c("Sample type", "Tissue", "Type", "Cancer", "Condition")) {
    if (cand %in% names(meta_ord) && is.character(meta_ord[[cand]])) {
      u <- unique(tolower(meta_ord[[cand]]))
      if (any(grepl("normal|healthy|control", u)) && any(grepl("tumor|tumour|cancer|pdac", u))) {
        sample_type_col <- cand
        break
      }
    }
  }
  if (is.null(sample_type_col) && "Sample type" %in% names(meta_ord)) sample_type_col <- "Sample type"

  ann_col <- data.frame(
    `PhenoMapR score` = score_norm,
    `Cell type` = factor(meta_ord[[ct_anno_col]]),
    `Prognostic group` = factor(meta_ord[[group_col]], levels = c("Most Adverse", "Other", "Most Favorable")),
    check.names = FALSE
  )
  if (!is.null(sample_type_col)) ann_col$`Sample type` <- factor(meta_ord[[sample_type_col]])
  rownames(ann_col) <- colnames(mat_scaled)

  pal_score <- colorRampPalette(c("#2166AC", "#F7F7F7", "#B2182B"))(100)
  pal_group <- c(`Most Adverse` = "#B2182B", Other = "#F7F7F7", `Most Favorable` = "#2166AC")
  pal_celltype <- PhenoMapR::get_celltype_palette(as.character(ann_col$`Cell type`))
  ann_colors <- list(
    `PhenoMapR score` = pal_score,
    `Cell type` = pal_celltype,
    `Prognostic group` = pal_group
  )
  if (!is.null(sample_type_col)) {
    st_lev <- levels(ann_col$`Sample type`)
    ann_colors$`Sample type` <- setNames(PhenoMapR::get_celltype_palette(st_lev), st_lev)
  }
  heatmap_colors <- if (requireNamespace("paletteer", quietly = TRUE)) {
    colorRampPalette(paletteer::paletteer_d("MexBrewer::Vendedora"))(100)
  } else {
    colorRampPalette(c("#2166AC", "#F7F7F7", "#B2182B"))(100)
  }
  if (requireNamespace("pheatmap", quietly = TRUE)) {
    pheatmap::pheatmap(
      mat_scaled,
      scale = "none",
      cluster_cols = FALSE,
      cluster_rows = FALSE,
      show_colnames = FALSE,
      annotation_col = ann_col,
      annotation_colors = ann_colors,
      color = heatmap_colors,
      breaks = seq(-3, 3, length.out = 101),
      main = "Top adverse & favorable marker genes (CRA001160)",
      fontsize_row = 8
    )
  } else {
    heatmap(mat_scaled, scale = "none", Colv = NA, col = heatmap_colors, labCol = FALSE, main = "Top marker genes (CRA001160)")
  }
}
```

PhenoMapR score is normalized to [-1, 1] so the bar is blue for negative, white at 0, and red for positive.

## Proportion of prognostic cells by sample and cell type

Stacked bar (top): proportion of Most Adverse and Most Favorable cells (5th percentile) per sample. Below: proportion within each sample and cell type.

```{r proportion-by-sample-celltype}
sample_col <- NULL
for (cand in c("Sample", "sample", "Patient", "patient", "Sample_ID", "orig.ident")) {
  if (cand %in% names(meta)) {
    sample_col <- cand
    break
  }
}
if (is.null(sample_col)) sample_col <- names(meta)[1]

meta_plot <- meta
meta_plot$sample <- meta_plot[[sample_col]]
meta_plot$cell_type <- meta_plot[[celltype_col]]
meta_plot$prognostic_grp <- meta_plot[[group_col]]
sample_lev <- levels(factor(meta[[sample_col]]))
meta_plot$sample <- factor(meta_plot[[sample_col]], levels = sample_lev)
n_per_sample <- setNames(as.numeric(table(meta[[sample_col]])[sample_lev]), sample_lev)
n_per_sample[is.na(n_per_sample)] <- 0

# Per-sample proportion of adverse and favorable (5th percentile) for stacked bar
meta_adv_fav <- meta_plot[meta_plot$prognostic_grp %in% c("Most Adverse", "Most Favorable"), ]
bar_df <- as.data.frame(table(sample = meta_adv_fav$sample, pg = meta_adv_fav$prognostic_grp, useNA = "no"))
bar_df <- bar_df[bar_df$pg %in% c("Most Adverse", "Most Favorable"), ]
bar_df$total <- n_per_sample[as.character(bar_df$sample)]
bar_df$proportion <- ifelse(bar_df$total > 0, bar_df$Freq / bar_df$total, 0)
fav <- bar_df[bar_df$pg == "Most Favorable", c("sample", "proportion")]
adv <- bar_df[bar_df$pg == "Most Adverse", c("sample", "proportion")]
names(fav)[2] <- "p_fav"
names(adv)[2] <- "p_adv"
bar_stack <- merge(data.frame(sample = sample_lev, stringsAsFactors = FALSE), fav, by = "sample", all.x = TRUE)
bar_stack <- merge(bar_stack, adv, by = "sample", all.x = TRUE)
bar_stack$p_fav[is.na(bar_stack$p_fav)] <- 0
bar_stack$p_adv[is.na(bar_stack$p_adv)] <- 0
bar_stack$sample <- factor(bar_stack$sample, levels = sample_lev)

meta_plot <- meta_plot[meta_plot$prognostic_grp %in% c("Most Adverse", "Most Favorable"), ]
meta_plot$cell_type <- factor(meta_plot$cell_type)

if (nrow(meta_plot) > 0) {
  counts <- as.data.frame(table(meta_plot$sample, meta_plot$cell_type, meta_plot$prognostic_grp, dnn = c("sample", "cell_type", "pg")))
  totals <- aggregate(counts$Freq, by = list(sample = counts$sample, pg = counts$pg), FUN = sum)
  names(totals)[3] <- "total"
  counts <- merge(counts, totals, by = c("sample", "pg"))
  counts$proportion <- counts$Freq / counts$total
  counts$proportion[counts$total == 0] <- 0
  counts$x_num <- as.numeric(counts$sample) + ifelse(counts$pg == "Most Adverse", -0.18, 0.18)

  # Stacked bar: adverse + favorable proportion per sample
  bar_long <- rbind(
    data.frame(sample = bar_stack$sample, pg = "Most Favorable", y_min = 0, y_max = bar_stack$p_fav),
    data.frame(sample = bar_stack$sample, pg = "Most Adverse", y_min = bar_stack$p_fav, y_max = bar_stack$p_fav + bar_stack$p_adv)
  )
  bar_long$pg <- factor(bar_long$pg, levels = c("Most Favorable", "Most Adverse"))
  p_bar <- ggplot(bar_long, aes(x = as.numeric(sample), fill = pg)) +
    geom_rect(aes(xmin = as.numeric(sample) - 0.4, xmax = as.numeric(sample) + 0.4, ymin = y_min, ymax = y_max)) +
    scale_fill_manual(values = c(`Most Adverse` = "#B2182B", `Most Favorable` = "#2166AC"), name = "Prognostic group") +
    scale_x_continuous(breaks = seq_along(sample_lev), labels = sample_lev) +
    scale_y_continuous(expand = c(0, 0)) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), axis.title.x = element_blank(), legend.position = "right") +
    labs(y = "Proportion (5th %ile)", title = "Adverse vs favorable per sample")

  p_dots <- ggplot(counts, aes(x = x_num, y = cell_type, size = proportion, color = pg)) +
    geom_point(alpha = 0.85) +
    scale_color_manual(values = c(`Most Adverse` = "#B2182B", `Most Favorable` = "#2166AC"), name = "Prognostic group") +
    scale_size_continuous(range = c(0, 8), name = "Proportion") +
    scale_x_continuous(breaks = seq_along(sample_lev), labels = sample_lev) +
    theme_minimal() +
    theme(panel.grid.major.y = element_line(color = "grey90"), axis.title = element_text(size = 10), axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "right") +
    labs(x = "Sample", y = "Cell type", title = "By cell type and sample (CRA001160)")

  if (requireNamespace("patchwork", quietly = TRUE)) {
    print(p_bar / p_dots + plot_layout(heights = c(1, 2))
  } else {
    print(p_bar)
    print(p_dots)
  }
} else {
  message("No Most Adverse or Most Favorable cells found; proportion plot skipped.")
}
```

## Conclusions

Here, we demonstrate that using PhenoMapR on single-cell RNAseq data in a pancreatic cancer dataset successfully identified cells known to be associated with disease (malignant and ductal type 1). These results agree with those of Jolasun et al., however our PhenoMapR results provide an increased level of granularity compared to other methods, since we retain absolute and rank-ordered information regarding all cell's phenotype association. We also nominate those most associated with favorable outcomes in PAAD, highlighting potential areas for additional therapeutic focus.

## References

- **CRA001160 (single-cell PDAC dataset)**: Peng J, Sun B-F, Chen C-Y, Zhou J-Y, Chen Y-S, et al. Single-cell RNA-seq highlights intra-tumoral heterogeneity and malignant progression in pancreatic ductal adenocarcinoma. *Cell Research* 29, 725–738 (2019). [https://doi.org/10.1038/s41422-019-0195-y](https://doi.org/10.1038/s41422-019-0195-y). Data: [GSA: CRA001160](https://ngdc.cncb.ac.cn/gsa/browse/CRA001160).

- **PRECOG 2.0**: Benard B, Lalgudi S, et al. PRECOG 2.0: an updated resource of pan-cancer gene-level prognostic meta-z scores. *Nucleic Acids Research*. 2026.

## Session Info

```{r session-info}
sessionInfo()
```
